---
title: "L04 Cluster Analysis"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: no
      smooth_scroll: yes
---

```{r setup, include=FALSE,message=FALSE,echo=FALSE}
#knitr::opts_knit$set(root.dir = "C:/Users/Declan/Desktop/Data/R/Cluster_analysis")
knitr::opts_knit$set(root.dir = "H:/Projects/10000/10682/TS/Scenario Manager/Cluster Analysis/Cleaned")

library(tidyverse)
library(lubridate)
library(gridExtra)
library(reshape2)
library(zoo)
library(plotly)
library(NbClust)
library(stringr)
```

# Read Data

Read data from Cluster Analysis Input Data.xlsx. From examining the file via Excel, we see that the data is located in the tab named "Database". We call that tab in the read_xlsx function.

```{r read data}
cluster_dat <- readxl::read_xlsx("Cluster Analysis Input Data.xlsx",sheet = 'Database')
```

Examine cluster_dat.

```{r structure of data}

str(cluster_dat)
```


First, we clean up the column headers by removing any potential trailing or leading spaces and replacing special characters in the strings with '_'.

```{r fix column names}

colnames(cluster_dat) <-trimws(colnames(cluster_dat))

colnames(cluster_dat) <-gsub('([[:punct:]])|\\s+','_',colnames(cluster_dat))

colnames(cluster_dat)
```

# DateTime

We look to fix the DateTime fields. The first five columns looks to be parsed versions of a DateTime. The 'Time Bin' looks to be the most complete DateTime variable. Examine further...

```{r check class of Time_Bin}
class(cluster_dat$Time_Bin)

cluster_dat$Time_Bin <- as.POSIXct(as.character(cluster_dat$Time_Bin),tz="US/Central")
```

```{r summary of Time_Bin}
summary(cluster_dat$Time_Bin)
table(is.na(cluster_dat$Time_Bin))
```

Everything seems to be in order. But we will keep an eye out for potential errors.

The last column is labelled week. Explore:

```{r summary of weeks}
summary(cluster_dat$Week)

```

Week 53 does not look right. Lets investigate further.

```{r check week 53}
cluster_dat[cluster_dat$Week==53,c(1:5,length(cluster_dat))]

```

Week 53 represents the last day of the year. We will ignore this field.

# Traffic Incidents

## Incidents
```{r incident summary}
summary(cluster_dat$Incident_Count)
```

```{r incident hist}
hist(cluster_dat$Incident_Count,breaks = 30)
```

Majority of days have no incidents.  

Explore the content of Incident Impact and Incident Count.

```{r incident plots}
summary(cluster_dat$Incident_Impact)

par(mfrow=c(2,2))

hist(cluster_dat$Incident_Impact,breaks=seq(0,40,1),
     xlab='Incident_Impact')

plot(cluster_dat$Incident_Impact[cluster_dat$Incident_Count!=0],cluster_dat$Incident_Count[cluster_dat$Incident_Count!=0],
     xlab="Incident_Impact",ylab='Incident_Count')

plot(cluster_dat$Time_Bin[cluster_dat$Incident_Count!=0],cluster_dat$Incident_Count[cluster_dat$Incident_Count!=0],
     xlab="Time",ylab='Incident_Count')

plot(cluster_dat$Density[cluster_dat$Incident_Count!=0],cluster_dat$Incident_Count[cluster_dat$Incident_Count!=0],
     xlab="Density",ylab='Incident_Count')
```

It appears that the incident count measures the amount on incidents in a specific time bin. The incident impact tracks how many impacts happened in a specific time_bin. This will allow us to determine how bad each time_bin is in terms of incidents.  

There are no easily detectable patterns when plotting incidents vs time but we do see a strong positive correlation between Density and Incidents counts. This makes sense as intuitively the more cars, the more incidents. 


We will focus on Incident Duration as a variable. We will add a variable 'Incident' that is a total incident minutes for each time bin.

```{r Incident Duration}

##create a total sum value for minutes of incidents in each time bin.
cluster_dat_incident_melt <- reshape2::melt(cluster_dat,id.vars="Time_Bin",measure.vars='Incident_Duration')%>%
  mutate(Hrs = as.numeric(format(value,"%H")),
         Mins = as.numeric(format(value,"%M")))%>%
  mutate(total = Hrs*60+Mins)%>%
  group_by(Time_Bin)%>%
  summarise(sum = sum(total))
  
cluster_dat$Incident <- cluster_dat_incident_melt$sum 

```


```{r incident severity bins, include=FALSE}
# We want to classify the Incident into severities. We will use the MNCMAT determination as a starting point for time bins:
# 
# * Fatal 180 mins
# * Incapacitating Injury 90 mins
# * Non-Incapacitating Injury 45 mins
# * Possible Injury 30 mins
# * Property Damage 30 mins
# * Unknown 30 mins
# 
# Our time bins will be:
# + = 0 mins,
# + 0 < x <= 30 mins,
# + 30 < x <= 45 mins,
# + 45 < x <= 90 mins,
# + 90 < x <= 180 mins,
# + 180 mins < x



# sev <-  data.frame(inc_sev=c('0 mins','30 mins','45 mins','90 mins','180 mins','> 180 mins'),
#                    symbol=c('circle', 'square', 'diamond','cross', 'x','triangle-down'))
# 
# 
# cluster_dat$inc_sev <- ifelse(cluster_dat$Incident==0,'0 mins',
#                               ifelse(cluster_dat$Incident>0&cluster_dat$Incident<30,'30 mins',
#                               ifelse(cluster_dat$Incident>30&cluster_dat$Incident<45,'45 mins',
#                               ifelse(cluster_dat$Incident>45&cluster_dat$Incident<90,'90 mins',
#                               ifelse(cluster_dat$Incident>90&cluster_dat$Incident<180,'180 mins',
#                               ifelse(cluster_dat$Incident>180,'> 180 mins','> 180 mins'))))))
# 
# cluster_dat <- merge(cluster_dat,sev,by='inc_sev')
```



## Roadwork

```{r summary roadwork count}
summary(cluster_dat$Roadwork_Count)
```

```{r table roadwork count}
table(cluster_dat$Roadwork_Count)
```

```{r roadwork plots}
par(mfrow=c(2,2))

hist(cluster_dat$Roadwork_Count[cluster_dat$Roadwork_Count!=0],xlab='Roadwork_Impact',main='Roadwork_Count')

plot(cluster_dat$Roadwork_Count[cluster_dat$Roadwork_Count!=0],cluster_dat$Roadwork_Impact[cluster_dat$Roadwork_Count!=0],
     xlab="Roadwork_Count",ylab='Roadwork_Impact')

plot(cluster_dat$Time_Bin[cluster_dat$Roadwork_Count!=0],cluster_dat$Roadwork_Count[cluster_dat$Roadwork_Count!=0],
     xlab="Time",ylab='Roadwork_Count')

plot(cluster_dat$Density[cluster_dat$Roadwork_Count!=0],cluster_dat$Roadwork_Count[cluster_dat$Roadwork_Count!=0],
     xlab="Density",ylab='Roadwork_Count')

```

```{r Roadwork Duration}

x <- as.numeric(format(cluster_dat$Roadwork_Duration,"%H"))
x <- x[x>0]
table(x)
```


Majority of days have no Roadwork. One really bad day with 7. No obvious outliers. 



# Metrics

Lets look at the Density, Volume, Speed variables. Again there are columns without names mixed in between these fields. We will ignore the unnamed variables.

### Date Time Fields

```{r create date time fields}
cluster_dat <- cluster_dat %>%
  mutate(Hour = format(Time_Bin,"%H"),
         Minute = format(Time_Bin,"%M"),
         Weekday = weekdays(cluster_dat$Time_Bin),
         Week = lubridate::week(cluster_dat$Time_Bin),
         Month = months(cluster_dat$Time_Bin),
         YearDay = yday(cluster_dat$Time_Bin))%>%
  filter(Hour == 14)%>%
  as.data.frame()




cluster_dat$Weekday <-  factor(cluster_dat$Weekday,levels=c("Sunday", "Monday", 
    "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
cluster_dat$Month <- factor(cluster_dat$Month,levels=c("January","February","March",
               "April","May","June","July","August","September",
               "October","November","December"),ordered=TRUE)
```


### Density

```{r density explore}
summary(cluster_dat$Density)

br_dens <- seq(0,50,2)
ranges_dens = paste(head(br_dens,-1), br_dens[-1], sep=" - ")
freq_dens <- hist(cluster_dat$Density, breaks=br_dens, include.lowest=TRUE, plot=FALSE)

freq_table_dens <-  data.frame(data.frame(range = ranges_dens, frequency = freq_dens$counts))
freq_table_dens[order(-freq_table_dens$frequency),]

plot(freq_dens)
```

The histogram is right skewed but no major outliers are present. There are also no 0 or NA values. 

We now plot the Density vs Time.

```{r weekday vs density}

ggplot(cluster_dat,aes(x=Weekday,y=Density,color=Weekday))+
  geom_violin()+
  geom_jitter(alpha=0.25)
```

Again, there do not seem to major outliers. There looks to be some patterns here. 

The Density is obviously lower on weekends compared to weekdays.




### Volume

```{r volume explore}
summary(cluster_dat$Volume)

br_vol <- seq(0,600000,20000)
ranges_vol = paste(head(br_vol,-1), br_vol[-1], sep=" - ")
freq_vol <- hist(cluster_dat$Volume, breaks=br_vol, include.lowest=TRUE, plot=FALSE)

freq_table_vol <-  data.frame(data.frame(range = ranges_vol, frequency = freq_vol$counts))
freq_table_vol[order(-freq_table_vol$frequency),]

plot(freq_vol)

```

It appears that the bin 40,000-60,000 contains the most records.





Again some interesting patterns but no major outliers. 

We move on to speed....


### Speed

We will explore the Speed variable. 

```{r speed explore}

summary(cluster_dat$Speed)

br_speed <- seq(30,100,2)
ranges_speed = paste(head(br_speed,-1), br_speed[-1], sep=" - ")
freq_speed <- hist(cluster_dat$Speed, breaks=br_speed, include.lowest=TRUE, plot=FALSE)

freq_table_speed <-  data.frame(data.frame(range = ranges_speed, frequency = freq_speed$counts))
freq_table_speed[order(-freq_table_speed$frequency),]

plot(freq_speed)


```

The speed plot looks more normal in shape than the previous distributions but does appear to have a left skew. Again there are no major outliers to be worried about nor are there are any 0 or NAs.

Lets plot speed vs time

```{r plot weekday vs volume}

means <-  aggregate(Speed~Weekday,cluster_dat,mean)
ggplot(cluster_dat,aes(x=Weekday,y=Speed))+
  geom_violin(aes(color=Weekday))+
  geom_jitter(aes(color=Weekday),alpha=0.25)+
  stat_summary(fun.y=mean, colour="#990000", geom="point", 
               shape=18, size=3,show_guide = FALSE)+
  geom_text(data=means, aes(label = paste('Mean:\n',round(Speed,2),sep=''), y = Speed + 2.5),size=3,fontface='bold')+
  ylab('Speed mph')+
  theme(legend.position="none")
```


```{r plot month vs volume}

means <-  aggregate(Speed~Month,cluster_dat,mean)
ggplot(cluster_dat,aes(x=Month,y=Speed))+
  geom_violin(aes(color=Month))+
  geom_jitter(aes(color=Month),alpha=0.25)+
  stat_summary(fun.y=mean, colour="#990000", geom="point", 
               shape=18, size=3,show_guide = FALSE)+
  geom_text(data=means, aes(label = paste('Mean:\n',round(Speed,2),sep=''), y = Speed + 2.5),color='darkred',size=3,fontface='bold')+
  ylab('Speed mph')+
  theme(legend.position="none")
```

Looks like we have some reduced speeds in the Summer months, maybe due to construction activities. 
There are also some slower days in the winter months which could very easily be due to bad weather. 
But there are no noticeable patterns. 

```{r speed vs week}
ggplot(cluster_dat,aes(x=Week,y=Speed,color=Weekday))+
  geom_point(aplha=0.25)+
  geom_smooth()
```

Again we are seeing a pattern with reduced speeds on weekends. 

Lets look at public holidays:

```{r create holiday list}
hols_2017 <- as.POSIXct(c('2017-01-02 14:00','2017-05-29 14:00','2017-07-03 14:00','2017-07-04 14:00','2017-09-04 14:00','2017-11-23 14:00','2017-11-24 14:00','2017-12-25 14:00'))
table(cluster_dat$Time_Bin %in% hols_2017)

cluster_dat$hols_2017 <- cluster_dat$Time_Bin %in% hols_2017

g <- ggplot(cluster_dat,aes(x=Weekday,y=Density))+
  geom_point(aes(alpha=0.25,color=hols_2017))
plotly::ggplotly(g)

s <- ggplot(cluster_dat,aes(x=Weekday,y=Speed))+
  geom_point(aes(alpha=0.25,color=hols_2017))
plotly::ggplotly(s)
```

Public holidays seem to be the weekdays with highest speeds.



### Weather

We need to generate a total minute count for weather events in each time bin. 

```{r add cumulative minutes for weather}
##create a total sum value for minutes of weather events in each time bin.
cluster_dat_melt_RS <- reshape2::melt(cluster_dat,id.vars="Time_Bin",measure.vars=c("Light_Rain","Light_Snow","Moderate_Rain","Moderate_Snow","Heavy_Snow","Heavy_Rain"))%>%
  mutate(Hrs = as.numeric(format(value,"%H")),
         Mins = as.numeric(format(value,"%M")),
         Percip = str_sub(variable,-4))%>%
  mutate(total = Hrs*60+Mins)%>%
  group_by(Time_Bin,Percip)%>%
  summarise(sum = sum(total))%>%
  dcast(Time_Bin~Percip,value.var = 'sum')
  
cluster_dat_melt_weather <- reshape2::melt(cluster_dat,id.vars="Time_Bin",measure.vars=colnames(cluster_dat)[11:16])%>%
  mutate(Hrs = as.numeric(format(value,"%H")),
         Mins = as.numeric(format(value,"%M")))%>%
  mutate(total = Hrs*60+Mins)%>%
  group_by(Time_Bin)%>%
  summarise(sum = sum(total))
  
cluster_dat$weather <- cluster_dat_melt_weather$sum

cluster_dat$Rain <- cluster_dat_melt_RS$Rain
cluster_dat$Snow <- cluster_dat_melt_RS$Snow

rm(cluster_dat_melt_RS,cluster_dat_melt_weather)
```

Now we filter out the weekends. We already determined that they will have their own cluster. 

## Filter Weekends

```{r filter midweek}

##filter out the weekends
'%ni%' <- Negate('%in%')
cluster_dat_midweek <- cluster_dat %>% 
  filter(Weekday %ni% c('Saturday','Sunday'))

## K Means Clustering {.tabset .tabset-fade}


```

# Clustering {.tabset .tabset-fade}

## Day of Year vs Speed

```{r Nbclust day~year}

nc <- NbClust(cluster_dat_midweek%>%select(Speed,YearDay,Rain,Snow,Incident)%>%scale(), min.nc=2, max.nc=15, method="kmeans")

```

We run the kmeans function with 5 centers and create a variable for the cluster designation. We will add this to our dataframe to facilitate plotting.

```{r run kmeans day~year}
set.seed(250)
#run kmeans
speed_cluster <-  kmeans(cluster_dat_midweek%>%select(Speed,YearDay,Rain,Snow,Incident)%>%scale(),centers = 5,nstart = 25)

##add cluster to fitlered dataframe
cluster_dat_midweek$cluster <- as.factor(speed_cluster$cluster)
```


We will look at some plots to display the clusters.

2D Plot Year Day ~ Speed

A standard 2d plot.

```{r 2d plots day~year}
plotly::plot_ly(cluster_dat_midweek,x=~YearDay,y=~Speed,color=~cluster,type='scatter',mode='markers',text=~paste('Time_Bin:', format(Time_Bin,"%Y-%m-%d")))
```

The 2d plot does not do a good job of showing where the clusters split.


## Year Day ~ Speed ~ Weather

```{r Nbclust day~year~weather}
set.seed(250)
nc <- NbClust(cluster_dat_midweek%>%select(Speed,YearDay,weather)%>%scale(), min.nc=2, max.nc=15, method="kmeans")

```

We run the kmeans function with 5 centers and create a variable for the cluster designation. We will add this to our dataframe to facilitate plotting.

```{r run kmeans day~year~weather}
set.seed(250)
#run kmeans
speed_cluster <-  kmeans(cluster_dat_midweek%>%select(Speed,YearDay,weather)%>%scale(),centers = 5,nstart = 25)

##add cluster to fitlered dataframe
cluster_dat_midweek$cluster <- as.factor(speed_cluster$cluster)
```


We will look at some plots to display the clusters.

2D Plot Year Day ~ Speed

First a simple 2d plot

```{r 2d plots day~year~weather}
plotly::plot_ly(cluster_dat_midweek,x=~YearDay,y=~Speed,color=~cluster,type='scatter',mode='markers',text=~paste('Time_Bin:', format(Time_Bin,"%Y-%m-%d")))
```

The 2d plot does not do a good job of showing where the clusters split.

3D Plot Year Day ~ Speed ~ Weather

we will try a 3d plot with weather on the z axis.

```{r 3d day~year~weather}
#3d plot to better show

plotly::plot_ly(cluster_dat_midweek,x=~YearDay,y=~Speed,z=~weather,
                color=~cluster, hoverinfo='text',
                text = ~paste('Speed:', round(Speed,2), '<br>Weather mins:', weather, '<br>Time_Bin:', format(Time_Bin,"%Y-%m-%d")))%>% 
  add_markers()%>%
    layout(scene = list(xaxis = list(title = 'Day of Year',
                                     range = c(0,max(cluster_dat_midweek$YearDay)),autorange=F),
                     yaxis = list(title = 'Speed MPH',
                                  range = c(min(cluster_dat_midweek$Speed),
                                            max(cluster_dat_midweek$Speed))),
                     zaxis = list(title = 'Weather mins',
                                  range = c(min(cluster_dat_midweek$weather),
                                            max(cluster_dat_midweek$weather)))))



```

Much better.



## Year Day ~ Speed ~ Weather ~ Incident

```{r Nbclust day~year~weather~incident}
set.seed(250)
nc <- NbClust(cluster_dat_midweek%>%select(Speed,YearDay,weather, Incident)%>%scale(), min.nc=2, max.nc=15, method="kmeans")

```

We run the kmeans function with 5 centers and create a variable for the cluster designation. We will add this to our dataframe to facilitate plotting.

```{r run kmeans day~year~weather~incident}
set.seed(250)
#run kmeans
speed_cluster <-  kmeans(cluster_dat_midweek%>%select(Speed,YearDay,weather, Incident)%>%scale(),centers = 5,nstart = 25)

##add cluster to fitlered dataframe
cluster_dat_midweek$cluster <- as.factor(speed_cluster$cluster)
```


We will look at some plots to display the clusters.


```{r 3d day~year~weather~incident}
#3d plot to better show

plotly::plot_ly(cluster_dat_midweek,x=~YearDay,y=~Speed,z=~weather,
                color=~cluster, hoverinfo='text',
                text = ~paste('Speed:', round(Speed,2), '<br>Weather mins:', weather, '<br>Time_Bin:', format(Time_Bin,"%Y-%m-%d")))%>% 
  add_markers()%>%
    layout(scene = list(xaxis = list(title = 'Day of Year',
                                     range = c(0,max(cluster_dat_midweek$YearDay)),autorange=F),
                     yaxis = list(title = 'Speed MPH',
                                  range = c(min(cluster_dat_midweek$Speed),
                                            max(cluster_dat_midweek$Speed))),
                     zaxis = list(title = 'Weather mins',
                                  range = c(min(cluster_dat_midweek$weather),
                                            max(cluster_dat_midweek$weather)))))



```

Much better.


3D Plot Year Day ~ Speed, Incidents

Now lets look at Incidents on the z axis.

```{r 3d plot day~year~weather~incident}
#3d plot to better show

plotly::plot_ly(cluster_dat_midweek,x=~YearDay,y=~Speed,z=~Incident,
                color=~cluster, hoverinfo='text',
                text = ~paste('Speed:', round(Speed,2), '<br>Incidents mins:', Incident, '<br>Time_Bin:', format(Time_Bin,"%Y-%m-%d")))%>% 
  add_markers()%>%
    layout(scene = list(xaxis = list(title = 'Day of Year',
                                     range = c(0,max(cluster_dat_midweek$YearDay)),autorange=F),
                     yaxis = list(title = 'Speed MPH',
                                  range = c(min(cluster_dat_midweek$Speed),
                                            max(cluster_dat_midweek$Speed))),
                     zaxis = list(title = 'Incident mins',
                                  range = c(min(cluster_dat_midweek$Incident),
                                            max(cluster_dat_midweek$Incident)))))

```



## Year Day ~ Speed ~Rain ~ Snow ~ Incident

```{r Nbclust day~year~rain~snow~incident}
set.seed(250)
nc <- NbClust(cluster_dat_midweek%>%select(Speed,YearDay,Rain, Snow, Incident)%>%scale(), min.nc=2, max.nc=15, method="kmeans")

```

We run the kmeans function with 5 centers and create a variable for the cluster designation. We will add this to our dataframe to facilitate plotting.

```{r run kmeans day~year~rain~snow~incident}
set.seed(250)
#run kmeans
speed_cluster <-  kmeans(cluster_dat_midweek%>%select(Speed,YearDay,Rain,Snow, Incident)%>%scale(),centers = 5,nstart = 25)

##add cluster to fitlered dataframe
cluster_dat_midweek$cluster <- as.factor(speed_cluster$cluster)
```


We will look at some plots to display the clusters.


```{r 3d day~year~rain~snow~incident}
#3d plot to better show

plotly::plot_ly(cluster_dat_midweek,x=~YearDay,y=~Speed,z=~Snow,
                color=~cluster, hoverinfo='text',
                text = ~paste('Speed:', round(Speed,2), '<br>Weather mins:', weather, '<br>Time_Bin:', format(Time_Bin,"%Y-%m-%d")))%>% 
  add_markers()%>%
    layout(scene = list(xaxis = list(title = 'Day of Year',
                                     range = c(0,max(cluster_dat_midweek$YearDay)),autorange=F),
                     yaxis = list(title = 'Speed MPH',
                                  range = c(min(cluster_dat_midweek$Speed),
                                            max(cluster_dat_midweek$Speed))),
                     zaxis = list(title = 'Snow mins',
                                  range = c(min(cluster_dat_midweek$Snow),
                                            max(cluster_dat_midweek$Snow)))))



```

Much better.


3D Plot Year Day ~ Speed, Incidents

Now lets look at Incidents on the z axis.

```{r 3d plot day~year~rain~snow~incident}
#3d plot to better show

plotly::plot_ly(cluster_dat_midweek,x=~YearDay,y=~Speed,z=~Incident,
                color=~cluster, hoverinfo='text',
                text = ~paste('Speed:', round(Speed,2), '<br>Incidents mins:', Incident, '<br>Time_Bin:', format(Time_Bin,"%Y-%m-%d")))%>% 
  add_markers()%>%
    layout(scene = list(xaxis = list(title = 'Day of Year',
                                     range = c(0,max(cluster_dat_midweek$YearDay)),autorange=F),
                     yaxis = list(title = 'Speed MPH',
                                  range = c(min(cluster_dat_midweek$Speed),
                                            max(cluster_dat_midweek$Speed))),
                     zaxis = list(title = 'Incident mins',
                                  range = c(min(cluster_dat_midweek$Incident),
                                            max(cluster_dat_midweek$Incident)))))

```



